---
title: "Identify population centers i"
---

# Data
+ GHS-WUP-MTUC R2025A from the European Commission JRC from the European Commission. It provides more than 10,000 globally harmonized urban centers
  + https://human-settlement.emergency.copernicus.eu/ghs_wup_mtuc_r2025a.php

# Process
I define population center $i$ as a GHSL Urban Centre, represented spatially by its population-weighted centroid.

Specifically, I use the Urban Centres identified in the GHS-WUP-MTUC R2025A dataset. An Urban Centre is defined using a globally harmonized Degree of Urbanisation framework rather than country-specific administrative boundaries. It consists of contiguous high-density grid cells with a population density of at least 1,500 persons per km$^2$ and a total population of at least 50,000.

Thus, the 50,000-person threshold is not an additional sample restriction imposed in this analysis; it is inherent in the GHSL definition of an Urban Centre. Using this definition provides a spatially consistent set of population centers across countries and avoids differences in national definitions of cities or municipalities.

For each population center $i$, I represent its location by the population-weighted centroid,

$$
\mathbf{x}_i =
(\mathrm{lon}_i^{PW}, \mathrm{lat}_i^{PW}),
$$

where $\mathrm{lon}_i^{PW}$ and $\mathrm{lat}_i^{PW}$ denote the longitude and latitude of the population-weighted centroid, respectively. This location is used as the representative municipal water-demand point in the subsequent analysis of surface-water access and conveyance costs.


```{r}

# /*===========================================*/
#'=  Downlaod the data (One time) =
# /*===========================================*/

# Directory
dir_data <- here("Data/Raw/GHS_WUP_MTUC_R2025A")

dir.create(
  dir_data,
  recursive = TRUE,
  showWarnings = FALSE
)

# Download URL
url_stats <- 
  paste0(
    "https://jeodpp.jrc.ec.europa.eu/ftp/jrc-opendata/GHSL/",
    "GHS_WUP_MTUC_GLOBE_R2025A/V1-1/",
    "GHS_WUP_MTUC_GLOBE_R2025A_V1_1_statistics.zip"
  )

# Local file
file_zip <- file.path(dir_data, "GHS_WUP_MTUC_GLOBE_R2025A_V1_1_statistics.zip")

download.file(
  url = url_stats,
  destfile = file_zip,
  mode = "wb"
)

# Unzip the downloaded file
dir_unzip <- file.path(
  dir_data,
  "GHS_WUP_MTUC_GLOBE_R2025A_V1_1_statistics"
)

dir.create(dir_unzip, showWarnings = FALSE)

unzip(file_zip, exdir = dir_unzip)
```


# Data Preparation

```{r}
# /*===========================================*/
#'=  Read the Excel file =
# /*===========================================*/
file_xlsx <- 
  list.files(
    here("Data/Raw/GHS_WUP_MTUC_R2025A/GHS_WUP_MTUC_GLOBE_R2025A_V1_1_statistics"),
    pattern = "\\.xlsx$",
    recursive = TRUE,
    full.names = TRUE
  )

# file_xlsx

# Read the data from the Excel file
uc_raw <- 
  rio::import(
    file_xlsx[1], 
    which = "UC_STATS",
    setclass = "data.table") %>%
  as.data.table()

names(uc_raw)
```


```{r}
# /*===========================================*/
#'=  Identify the 2025 population centers =
# /*===========================================*/
uc_2025 <- 
  uc_raw[
    Year == 2025,
    .(
      urban_id     = ID_UC_G0,
      urban_name   = UCname,
      country_id   = UNLocID,
      country_name = UNLocName,
      population   = POP,
      area_km2     = AREA_km2,
      lon          = Lon,
      lat          = Lat,
      capital      = CapitalFlag,
      size_class   = size_class_epoch,
      plausibility = Plausibility
    )
  ]

nrow(uc_2025) #12,140
uniqueN(uc_2025$urban_id) #12,140
uc_2025[, .N, by = urban_id][N > 1] # So, there are no duplicate urban_id values in the dataset.

saveRDS(uc_2025, here("Data/Processed/ghsl_urban_centres_2025.rds"))

uc_2025_sf <- st_as_sf(
  uc_2025,
  coords = c("lon", "lat"),
  crs = 4326,
  remove = FALSE
)

st_write(
  uc_2025_sf,
  here("Data/Processed/ghsl_urban_centers_2025.gpkg")
)

#/*--------------------------------*/
#' ## template
#/*--------------------------------*/
# uc_2025_template <-
#   uc_2025[
#     , .(
#       urban_id,
#       urban_name,
#       country_id,
#       country_name,
#       population,
#       lon, lat
#     )
#   ]

# fwrite(
#   uc_2025_template,
#   here("Data/Processed/ghsl_urban_centers_2025_template.csv"),
#   bom = TRUE
# )

# uc_2025_template[urban_id == 747]
```

# Plot

```{r}
uc_2025_sf <- 
  st_as_sf(
    uc_2025,
    coords = c("lon", "lat"),
    crs = 4326,
    remove = FALSE
  )

world_bd <- st_read(here("Data/Raw/boundary/z_ee_r250_correspondence.gpkg")) %>%
  st_transform(crs = 4326)

ggplot() + 
  geom_sf(data = world_bd, fill = "gray90", color = "gray50") +
  geom_sf(data = uc_2025_sf, aes(size = population), color = "blue", alpha = 0.5)
```




