---
title: "Identify candidate river, lake, and reservoir sources"
---

This file creates the broad surface-water candidate list for `4_2_surface_water_shortlist.qmd`. It does not set withdrawal capacities.  Candidate intakes inside mapped protected-area polygons are excluded before ranking. The screen is a study assumption about geographic access, not a determination of water rights. The proposed intake is the closest point on each river reach or lake polygon to the city; a different permitted intake on the same source is not searched for here.

Download the polygon layer of the World Database on Protected Areas from [Protected Planet](https://www.protectedplanet.net/en/search-areas). Use the protected-area (PA) polygons, including terrestrial and inland-water sites.  The combined WDPCA download can also include OECMs and point-only records; this script excludes OECMs and cannot screen areas known only as points.  The protected-area path and polygon layer below refer to the September 2026 WDPA geodatabase in `Data/Raw/WDPA_Sep2026_Public`. Update them only if that dataset is moved or replaced. Keep a record of the downloaded data version.

River reaches must have at least 0.5 m3/s modeled mean annual naturalized flow. The lowest-month flow is retained as an attribute, but it does not exclude reaches from this broad candidate file. Step 4_2 uses both mean annual and lowest-month flow to select complementary river shortlists.  Step 5 calculates source capacity from mean annual flow.  LakeATLAS starts at about 0.1 km2. There is no additional lake-size cutoff in this baseline because we have not checked how a larger threshold would affect known municipal sources. Lake candidates are diversified across nearby sources, high average outlet flow, large area, and reservoir volume. `Vol_res` is a storage descriptor; it is not annual supply.

```{r}
#| label: setup

library(data.table)
library(sf)
library(here)
library(parallel)

city_file <- here(
  "Data/Analysis/ghsl_urban_centres_2019_with_muni_demand.csv"
)

river_gdb <- here(
  "Data/Raw/RiverATLAS/RiverATLAS_v10.gdb"
  )

lake_gdb <- here(
  "Data/raw/LakeATLAS/LakeATLAS_v10.gdb"
  )

protected_area_path <- here(
  "Data/Raw/WDPA_Sep2026_Public/WDPA_Sep2026_Public.gdb"
)

protected_area_layer <- "WDPA_poly_Sep2026"

max_distance_km <- 100
min_river_meanflow_m3s <- 0.5

# Baseline reproduces the approximate lower area limit of LakeATLAS.
# Raise this threshold only after checking known municipal lake intakes.
min_lake_area_km2 <- 0.1

# At most 20 lakes per city, with complementary ranking criteria.
n_lake_nearest <- 10L
n_lake_highflow <- 5L
n_lake_largest_area <- 3L
n_lake_reservoir <- 2L

tile_size_deg <- 5
n_cores <- 10L
protected_point_batch_size <- 25000L

required_inputs <- c(city_file, river_gdb, lake_gdb, protected_area_path)
missing_inputs <- required_inputs[!file.exists(required_inputs)]

if (length(missing_inputs)) {
  stop(
    "Missing inputs: ",
    paste(missing_inputs, collapse = ", "),
    ". Download the PA polygon dataset and update protected_area_path."
  )
}

city <- fread(city_file)

stopifnot(all(c(
  "urban_id", "lon", "lat", "muni_withdrawal_screen_m3s"
) %in% names(city)))

city <- city[is.finite(lon) & is.finite(lat)]

stopifnot(!anyDuplicated(city$urban_id))

cities_sf <- st_as_sf(
  city,
  coords = c("lon", "lat"),
  crs = 4326,
  remove = FALSE
)

cities_sf$tile_id <- paste(
  floor((cities_sf$lon + 180) / tile_size_deg),
  floor((cities_sf$lat + 90) / tile_size_deg),
  sep = "_"
)

tile_ids <- unique(cities_sf$tile_id)

river_layers <- st_layers(river_gdb)$name
stopifnot(length(river_layers) == 1L)
river_layer <- river_layers[1L]

lake_layer <- "LakeATLAS_v10_pol"
stopifnot(lake_layer %in% st_layers(lake_gdb)$name)

pa_layers <- st_layers(protected_area_path)

if (is.null(protected_area_layer)) {
  polygon_layers <- pa_layers$name[
    vapply(
      pa_layers$geomtype,
      function(type) any(grepl("polygon", type, ignore.case = TRUE)),
      logical(1)
    )
  ]

  if (length(polygon_layers) != 1L) {
    stop(
      "Set protected_area_layer to the PA polygon layer. Found: ",
      paste(polygon_layers, collapse = ", ")
    )
  }

  protected_area_layer <- polygon_layers[1L]
}

stopifnot(protected_area_layer %in% pa_layers$name)

pa_layer_index <- match(protected_area_layer, pa_layers$name)
pa_crs <- pa_layers$crs[[pa_layer_index]]

if (is.na(pa_crs)) {
  stop("The protected-area polygon layer needs a defined CRS")
}

read_protected_tile <- function(search_bbox) {
  # The bbox must use the layer's CRS for GDAL's spatial filter.
  filter_bbox <- st_transform(
    search_bbox,
    pa_crs
  )

  pa <- st_read(
    protected_area_path,
    layer = protected_area_layer,
    wkt_filter = st_as_text(filter_bbox),
    quiet = TRUE
  )

  if (nrow(pa) == 0L) {
    return(st_sfc(crs = 4326))
  }

  # WDPCA uses SITE_TYPE = PA or OECM; WDPA-only files may omit it.
  pa_type_column <- names(pa)[
    tolower(names(pa)) == "site_type"
  ]

  pa_status_column <- names(pa)[
    tolower(names(pa)) == "status"
  ]

  if (length(pa_type_column)) {
    pa <- pa[
      !is.na(pa[[pa_type_column]]) &
        toupper(as.character(pa[[pa_type_column]])) == "PA",
    ]
  }

  if (length(pa_status_column)) {
    # Proposed sites are excluded. Sites with unknown status remain.
    pa <- pa[
      is.na(pa[[pa_status_column]]) |
        toupper(as.character(pa[[pa_status_column]])) != "PROPOSED",
    ]
  }

  if (nrow(pa) == 0L) {
    return(st_sfc(crs = 4326))
  }

  # Repair WDPA polygons in a local projected CRS. Spherical s2 validation
  # can reject self-crossing WDPA rings even when a projected GEOS repair works.
  box <- st_bbox(search_bbox)
  tile_crs <- st_crs(sprintf(
    paste(
      "+proj=aeqd",
      "+lat_0=%.8f",
      "+lon_0=%.8f",
      "+datum=WGS84",
      "+units=m",
      "+no_defs"
    ),
    mean(c(box["ymin"], box["ymax"])),
    mean(c(box["xmin"], box["xmax"]))
  ))

  pa <- st_transform(pa, tile_crs)
  pa <- st_make_valid(pa)

  geometry_types <- as.character(st_geometry_type(pa))
  if (any(!geometry_types %in% c("POLYGON", "MULTIPOLYGON"))) {
    pa <- st_collection_extract(pa, "POLYGON", warn = FALSE)
  }

  st_geometry(pa)
}

tile_search_bbox <- function(cities_tile) {
  city_buffers <- st_buffer(
    st_geometry(cities_tile),
    dist = max_distance_km * 1000
  )

  st_as_sfc(st_bbox(city_buffers))
}

source_point_coordinates <- function(city_point, sources) {
  city_points <- st_geometry(city_point)[
    rep(1L, nrow(sources))
  ]

  nearest_lines <- st_nearest_points(
    city_points,
    st_geometry(sources),
    pairwise = TRUE
  )

  t(vapply(
    nearest_lines,
    function(line) {
      coords <- st_coordinates(line)
      coords[nrow(coords), c("X", "Y")]
    },
    numeric(2)
  ))
}

keep_unprotected_intakes <- function(candidates, protected_polygons) {
  if (nrow(candidates) == 0L || length(protected_polygons) == 0L) {
    return(candidates)
  }

  inside <- rep(FALSE, nrow(candidates))
  batch_starts <- seq.int(
    1L,
    nrow(candidates),
    by = protected_point_batch_size
  )

  for (start in batch_starts) {
    rows <- seq.int(
      start,
      min(start + protected_point_batch_size - 1L, nrow(candidates))
    )

    intakes <- st_as_sf(
      candidates[rows, .(source_lon, source_lat)],
      coords = c("source_lon", "source_lat"),
      crs = 4326
    )
    intakes <- st_transform(intakes, st_crs(protected_polygons))

    inside[rows] <- lengths(
      st_intersects(intakes, protected_polygons)
    ) > 0L
  }

  candidates[!inside]
}
```

# River candidates

All eligible reaches within 100 km remain available for the network-based shortlist in step 4_2. Each source point is tested against protected areas.

```{r}
#| label: identify-river-candidates

process_river_tile <- function(tile_id) {
  cities_tile <- cities_sf[cities_sf$tile_id == tile_id, ]
  search_bbox <- tile_search_bbox(cities_tile)

  rivers <- st_read(
    river_gdb,
    layer = river_layer,
    wkt_filter = st_as_text(search_bbox),
    quiet = TRUE
  )

  if (nrow(rivers) == 0L) {
    return(NULL)
  }

  required_fields <- c("HYRIV_ID", "dis_m3_pyr", "dis_m3_pmn")
  stopifnot(all(required_fields %in% names(rivers)))

  rivers <- rivers[
    is.finite(rivers$dis_m3_pyr) &
      rivers$dis_m3_pyr >= min_river_meanflow_m3s,
  ]

  if (nrow(rivers) == 0L) {
    return(NULL)
  }

  nearby <- st_is_within_distance(
    cities_tile,
    rivers,
    dist = max_distance_km * 1000
  )

  candidates_tile <- rbindlist(lapply(seq_len(nrow(cities_tile)), function(j) {
    idx <- nearby[[j]]
    if (length(idx) == 0L) return(NULL)

    distance_m <- as.numeric(st_distance(cities_tile[j, ], rivers[idx, ]))
    coords <- source_point_coordinates(cities_tile[j, ], rivers[idx, ])

    ord <- order(distance_m, rivers$HYRIV_ID[idx])
    idx <- idx[ord]
    distance_m <- distance_m[ord]
    coords <- coords[ord, , drop = FALSE]

    if (anyDuplicated(rivers$HYRIV_ID[idx])) {
      stop("Duplicate HYRIV_ID in RiverATLAS spatial query")
    }

    data.table(
      urban_id = cities_tile$urban_id[j],
      river_id = rivers$HYRIV_ID[idx],
      distance_km = distance_m / 1000,
      source_lon = coords[, 1L],
      source_lat = coords[, 2L],
      river_discharge_m3s = rivers$dis_m3_pyr[idx],
      river_lowflow_m3s = rivers$dis_m3_pmn[idx],
      muni_demand_m3s = cities_tile$muni_withdrawal_screen_m3s[j]
    )
  }), use.names = TRUE, fill = TRUE)

  if (nrow(candidates_tile) == 0L) return(NULL)

  # Check intake points in batches across the tile, rather than
  # building a new protected-area index for every city.
  protected_tile <- read_protected_tile(search_bbox)
  candidates_tile <- keep_unprotected_intakes(
    candidates_tile,
    protected_tile
  )

  if (nrow(candidates_tile) == 0L) return(NULL)

  candidates_tile[, candidate_rank := seq_len(.N), by = urban_id]
  candidates_tile
}

river_parts <- mclapply(
  tile_ids,
  process_river_tile,
  mc.cores = n_cores
)

river_candidates <- rbindlist(
  river_parts,
  use.names = TRUE,
  fill = TRUE
)

stopifnot(!anyDuplicated(
  river_candidates[, .(urban_id, river_id)]
))

saveRDS(
  river_candidates,
  here("Data/Processed/0_river_candidates_raw.rds")
)
```

# Lake and reservoir candidates

The default area screen does not remove LakeATLAS's smallest lakes. A higher
cutoff requires a sensitivity check against known municipal intakes. A lake
with a recorded `Vol_res > 0` remains eligible even if it falls below the
chosen area threshold. A positive modeled average outlet flow is required
for the subsequent flow-based cost scenario.
Keep up to 20 distinct candidates per city, selected across four rankings.
For a partially protected lake, this checks the proposed intake point; it does
not assert that every point on the shoreline has the same protected status.

```{r}
#| label: identify-lake-candidates

process_lake_tile <- function(tile_id) {
  cities_tile <- cities_sf[cities_sf$tile_id == tile_id, ]
  search_bbox <- tile_search_bbox(cities_tile)

  lakes <- st_read(
    lake_gdb,
    layer = lake_layer,
    wkt_filter = st_as_text(search_bbox),
    quiet = TRUE
  )

  if (nrow(lakes) == 0L) {
    return(NULL)
  }

  required_fields <- c(
    "Hylak_id", "Lake_type", "Grand_id", "Lake_area",
    "Vol_total", "Vol_res", "dis_m3_pyr", "dis_m3_pmn",
    "Elevation"
  )
  stopifnot(all(required_fields %in% names(lakes)))

  eligible_size <- (
    is.finite(lakes$Lake_area) &
      lakes$Lake_area >= min_lake_area_km2
  ) | (
    is.finite(lakes$Vol_res) &
      lakes$Vol_res > 0
  )

  lakes <- lakes[
    eligible_size &
      is.finite(lakes$dis_m3_pyr) &
      lakes$dis_m3_pyr > 0,
  ]

  if (nrow(lakes) == 0L) {
    return(NULL)
  }

  lakes <- st_make_valid(lakes)

  nearby <- st_is_within_distance(
    cities_tile,
    lakes,
    dist = max_distance_km * 1000
  )

  candidates_tile <- rbindlist(lapply(seq_len(nrow(cities_tile)), function(j) {
    idx <- nearby[[j]]
    if (length(idx) == 0L) return(NULL)

    distance_m <- as.numeric(st_distance(cities_tile[j, ], lakes[idx, ]))
    coords <- source_point_coordinates(cities_tile[j, ], lakes[idx, ])

    data.table(
      urban_id = cities_tile$urban_id[j],
      lake_id = lakes$Hylak_id[idx],
      distance_km = distance_m / 1000,
      source_lon = coords[, 1L],
      source_lat = coords[, 2L],
      lake_type = lakes$Lake_type[idx],
      grand_id = lakes$Grand_id[idx],
      lake_discharge_m3s = lakes$dis_m3_pyr[idx],
      lake_lowflow_m3s = lakes$dis_m3_pmn[idx],
      lake_area_km2 = lakes$Lake_area[idx],
      lake_volume_mcm = lakes$Vol_total[idx],
      reservoir_volume_mcm = lakes$Vol_res[idx],
      lake_elevation_m = lakes$Elevation[idx],
      muni_demand_m3s = cities_tile$muni_withdrawal_screen_m3s[j]
    )
  }), use.names = TRUE, fill = TRUE)

  if (nrow(candidates_tile) == 0L) return(NULL)

  protected_tile <- read_protected_tile(search_bbox)
  candidates_tile <- keep_unprotected_intakes(
    candidates_tile,
    protected_tile
  )

  if (nrow(candidates_tile) == 0L) return(NULL)

  # Rank only unprotected candidates. .I returns row numbers in the
  # combined tile table; selecting these rows keeps the original metadata.
  selected_rows <- candidates_tile[, {
    nearest <- head(
      order(distance_km, lake_id),
      n_lake_nearest
    )

    highflow <- head(
      order(-lake_discharge_m3s, distance_km),
      n_lake_highflow
    )

    largest_area <- head(
      order(-lake_area_km2, distance_km),
      n_lake_largest_area
    )

    reservoirs <- which(
      is.finite(reservoir_volume_mcm) &
        reservoir_volume_mcm > 0
    )

    large_reservoirs <- if (length(reservoirs)) {
      reservoirs[head(
        order(-reservoir_volume_mcm[reservoirs],
              distance_km[reservoirs]),
        n_lake_reservoir
      )]
    } else {
      integer()
    }

    chosen <- unique(c(
      nearest,
      highflow,
      largest_area,
      large_reservoirs
    ))

    .(row = .I[chosen])
  }, by = urban_id]$row

  candidates_tile <- candidates_tile[selected_rows]
  setorder(candidates_tile, urban_id, distance_km, lake_id)
  candidates_tile[, candidate_rank := seq_len(.N), by = urban_id]
  candidates_tile
}

lake_parts <- mclapply(
  tile_ids,
  process_lake_tile,
  mc.cores = n_cores
)

lake_candidates <- rbindlist(
  lake_parts,
  use.names = TRUE,
  fill = TRUE
)

stopifnot(!anyDuplicated(lake_candidates[, .(urban_id, lake_id)]))

lake_counts <- lake_candidates[, .N, by = urban_id]
stopifnot(all(
  lake_counts$N <= n_lake_nearest + n_lake_highflow +
    n_lake_largest_area + n_lake_reservoir
))

saveRDS(
  lake_candidates,
  here("Data/Processed/0_lake_candidates_raw.rds")
)
```

# Combine and validate

The combined file retains the metadata needed to audit lake and reservoir
selection and to compute mean-annual-flow capacity in step 5. Step 4_2 also
uses lowest-month flow for its additional shortlist. Changing the candidate
locations means the elevation and downstream cost steps must be rerun.

```{r}
#| label: combine-and-validate-and-save

river_candidates[, `:=`(
  source_type = "river",
  source_id = as.character(river_id),
  source_discharge_m3s = river_discharge_m3s,
  source_lowflow_m3s = river_lowflow_m3s,
  source_elevation_m = NA_real_
)]

lake_candidates[, `:=`(
  source_type = "lake",
  source_id = as.character(lake_id),
  source_discharge_m3s = lake_discharge_m3s,
  source_lowflow_m3s = lake_lowflow_m3s,
  source_elevation_m = lake_elevation_m
)]

common_columns <- c(
  "urban_id", "source_type", "source_id", "candidate_rank",
  "distance_km", "source_lon", "source_lat",
  "source_discharge_m3s", "source_lowflow_m3s",
  "source_elevation_m", "muni_demand_m3s"
)

lake_metadata <- c(
  "lake_type", "grand_id", "lake_area_km2",
  "lake_volume_mcm", "reservoir_volume_mcm"
)

river_part <- river_candidates[, ..common_columns]
river_part[, (lake_metadata) := NA_real_]

lake_part <- lake_candidates[,
  c(common_columns, lake_metadata),
  with = FALSE
]

surface_water_candidates <- rbindlist(
  list(river_part, lake_part),
  use.names = TRUE
)

stopifnot(
  !anyDuplicated(surface_water_candidates[
    , .(urban_id, source_type, source_id)
  ]),
  all(is.finite(surface_water_candidates$source_discharge_m3s)),
  all(surface_water_candidates[
    source_type == "river", source_discharge_m3s
  ] >= min_river_meanflow_m3s),
  all(surface_water_candidates$distance_km <= max_distance_km + 1e-8)
)

candidate_counts <- surface_water_candidates[
  , .N, by = .(source_type, urban_id)
]

print(candidate_counts[, .(
  cities = .N,
  candidates = sum(N),
  median_per_city = median(N),
  max_per_city = max(N)
), by = source_type])

print(surface_water_candidates[
  source_type == "lake",
  .N,
  by = lake_type
])

saveRDS(
  surface_water_candidates,
  here("Data/Processed/1_surface_water_candidates.rds")
)
```
