---
title: "Get and reuse elevation data for cities and surface-water intakes"
---


`surface_water_candidates` contains:

+ elevation for lakes and reservoirs: `source_elevation_m`

We need to get the (1) elevation for population centers `i` and (2) elevation for river reaches `s` 

# Data
+ FABDEM V1-2
  + 30m resolution (1 arc-second)
  + It is derived from Copernicus GLO-30 but specifically removes building and forest-height biases, so it is closer to bare-earth terrain elevation.
  + Its horizontal reference is WGS84 and vertical elevations are referenced to EGM2008.
  + It is much more suitable than a DSM for urban-centre centroids in particular.


```{r}
uc_2025 <- 
  readRDS(
    here("Data/Processed/ghsl_urban_centres_2025.rds")
  )

surface_water_candidates <- 
  readRDS(
    here("Data/Processed/1_surface_water_candidates_shortlist.rds")
  )

# range(uc_2025$lat, na.rm = TRUE)
# range(surface_water_candidates$source_lat, na.rm = TRUE)
```


```{r}
# Read the FABDEM tile index
fabdem_tiles <- 
  st_read(
    here("Data/Raw/FABDEM/FABDEM_v1-2_tiles.geojson"),
    quiet = TRUE
  )

# Create all locations wehre we need elevation =
fabdem_points <- 
  rbindlist(
    list(
      # Population centers i
      uc_2025[, .(point_type = "urban", lon, lat)], 
      # river reaches s
      surface_water_candidates[, .(point_type = "source", lon = source_lon, lat = source_lat)]
    )
  )

fabdem_points <- unique(fabdem_points[ is.finite(lon) & is.finite(lat) ])

# Verify that FABDEM covers all required points
fabdem_points_sf <- 
  st_as_sf(
    fabdem_points,
    coords = c("lon", "lat"),
    crs = 4326,
    remove = FALSE
  )

fabdem_covered <- lengths(st_intersects(fabdem_points_sf, fabdem_tiles)) > 0L
fabdem_uncovered <- fabdem_points_sf[!fabdem_covered, ]

table(fabdem_covered)
table(fabdem_uncovered$point_type)


if (nrow(fabdem_uncovered)) {
  nearest_tile <- st_nearest_feature(fabdem_uncovered, fabdem_tiles)

  fabdem_uncovered$distance_to_fabdem_m <- as.numeric(
    st_distance(
      fabdem_uncovered,
      fabdem_tiles[nearest_tile, ],
      by_element = TRUE
    )
  )

  summary(fabdem_uncovered$distance_to_fabdem_m)
}

fabdem_uncovered_dt <- as.data.table(st_drop_geometry(fabdem_uncovered))
```




# Reuse previously completed elevations by exact point type and coordinates

```{r}
fabdem_points_sf$elev_point_id <- seq_len(nrow(fabdem_points_sf))
fabdem_points_sf$fabdem_covered <- fabdem_covered

# Create a lookup table for the points
fabdem_points_lookup <- 
  as.data.table(
    st_drop_geometry(fabdem_points_sf)
  )
# This lookup will eventually hold the elevation for every unique urban/source corrdinate pair

# An elev_point_id is only a row number for this run. Do not use an old
# elev_point_id to match an old elevation to a new candidate list.
elevation_dir <- here("Data/Processed/FABDEM")
previous_master <- file.path(
  elevation_dir,
  "elevation_points_complete.rds"
)

# If the whole FABDEM folder was renamed during an earlier rerun, the
# completed master can also be read from that older folder.
if (!file.exists(previous_master)) {
  previous_master <- here(
    "Data/Processed/FABDEM_previous_candidates/elevation_points_complete.rds"
  )
}

cached_elevations <- data.table(
  point_type = character(),
  lon = numeric(),
  lat = numeric(),
  elevation_m = numeric(),
  elevation_dataset = character()
)

if (file.exists(previous_master)) {
  old <- as.data.table(readRDS(previous_master))
  needed <- names(cached_elevations)
  stopifnot(all(needed %in% names(old)))

  old <- old[
    is.finite(lon) & is.finite(lat) & is.finite(elevation_m) &
      elevation_dataset %in% c(
        "FABDEM_V1-2",
        "COPERNICUS_GLO30_2024_1"
      ),
    ..needed
  ]

  # A single point cannot have two different saved elevations.
  inconsistent <- old[, .(
    n_values = uniqueN(elevation_m),
    n_datasets = uniqueN(elevation_dataset)
  ), by = .(point_type, lon, lat)][
    n_values > 1L | n_datasets > 1L
  ]
  if (nrow(inconsistent)) {
    stop("Previous elevation master has conflicting values for a point")
  }

  cached_elevations <- unique(
    old,
    by = c("point_type", "lon", "lat")
  )
}

fabdem_points_lookup[, `:=`(
  cached_elevation_m = NA_real_,
  cached_elevation_dataset = NA_character_
)]

fabdem_points_lookup[
  cached_elevations,
  on = .(point_type, lon, lat),
  `:=`(
    cached_elevation_m = i.elevation_m,
    cached_elevation_dataset = i.elevation_dataset
  )
]

cat(
  "Points with saved elevation:",
  sum(is.finite(fabdem_points_lookup$cached_elevation_m)),
  "/", nrow(fabdem_points_lookup), "\n"
)

# Match each covered point to its actual FABDEM tile
covered_points_sf <- fabdem_points_sf[fabdem_covered, ]

tile_match <- st_intersects(
  covered_points_sf,
  fabdem_tiles
)

# Every covered point should intersect at least one tile
stopifnot(all(lengths(tile_match) >= 1L))

# If a point lies exactly on a tile boundary, use the first matching tile
tile_idx <- vapply(
  tile_match,
  function(x) x[1L],
  integer(1)
)

covered_points_sf$tile_name <- fabdem_tiles$tile_name[tile_idx]

covered_points_sf$file_name <- fabdem_tiles$file_name[tile_idx]

covered_points_sf$zipfile_name <- fabdem_tiles$zipfile_name[tile_idx]


# Check the attributes
head(st_drop_geometry(covered_points_sf)[, c("lon", "lat", "tile_name", "file_name", "zipfile_name") ])
```


```{r}

#/*==========================================================*/
#' ## 1. Prepare covered FABDEM points
#/*==========================================================*/

# covered_points_sf must already contain:
# elev_point_id
# point_type
# lon
# lat
# tile_name
# file_name
# zipfile_name

stopifnot(
  all(
    c(
      "elev_point_id",
      "point_type",
      "lon",
      "lat",
      "tile_name",
      "file_name",
      "zipfile_name"
    ) %in% names(covered_points_sf)
  )
)

fabdem_points_covered <-
  as.data.table(
    st_drop_geometry(covered_points_sf)
  )[
    !elev_point_id %in%
      fabdem_points_lookup[
        is.finite(cached_elevation_m),
        elev_point_id
      ],
    .(
      elev_point_id,
      point_type,
      lon,
      lat,
      tile_name
    )
  ]


#/*==========================================================*/
#' ## 2. Construct required FABDEM tile table
#/*==========================================================*/

fabdem_required_tiles <-
  unique(
    as.data.table(
      st_drop_geometry(covered_points_sf)
    )[
      elev_point_id %in% fabdem_points_covered$elev_point_id,
      .(
        tile_name,
        file_name,
        zipfile_name
      )
    ]
  )

setorder(
  fabdem_required_tiles,
  tile_name
)


#/*==========================================================*/
#' ## 3. Normalize FABDEM mirror names
#/*==========================================================*/

# Hugging Face bucket
fabdem_base_url <-
  "https://huggingface.co/buckets/links-ads/fabdem/resolve/tiles"


# ---- Parent 10-degree folder ---- #

fabdem_required_tiles[
  ,
  parent_folder := sub(
    "\\.zip$",
    "",
    zipfile_name
  )
]


# Fix malformed southern/northern endpoint notation:
# e.g.
# S60W070-S-50W060 -> S60W070-S50W060

fabdem_required_tiles[
  ,
  parent_folder := gsub(
    "([NS])-([0-9]{2})",
    "\\1\\2",
    parent_folder
  )
]


# Prime meridian convention:
# W000 -> E000 when it is the eastern boundary

fabdem_required_tiles[
  ,
  parent_folder := sub(
    "-([NS][0-9]{2})W000_",
    "-\\1E000_",
    parent_folder
  )
]


# Antimeridian convention:
# E180 -> W180

fabdem_required_tiles[
  ,
  parent_folder := sub(
    "-([NS][0-9]{2})E180_",
    "-\\1W180_",
    parent_folder
  )
]


# ---- Individual 1-degree tile name ---- #
#
# Official index:
# N000E006
# S055W068
#
# Hugging Face mirror:
# N00E006
# S55W068

fabdem_required_tiles[,
  bucket_tile := sub(
    "^([NS])0([0-9]{2})([EW][0-9]{3})$",
    "\\1\\2\\3",
    tile_name
  )
]


fabdem_required_tiles[,
  mirror_file := paste0(bucket_tile, "_FABDEM_V1-2.tif") 
]

# ---- Final URL ---- #
fabdem_required_tiles[,
  url := paste0(fabdem_base_url, "/", parent_folder, "/", mirror_file)
]


#/*==========================================================*/
#' ## 4. Validate naming BEFORE any downloading
#/*==========================================================*/

# No malformed S-50 / N-50 style endpoints
stopifnot(
  !any(
    grepl(
      "[NS]-[0-9]{2}",
      fabdem_required_tiles$parent_folder
    )
  )
)

# No eastern boundary expressed as W000
stopifnot(
  !any(
    grepl(
      "-[NS][0-9]{2}W000_",
      fabdem_required_tiles$parent_folder
    )
  )
)

# No eastern boundary expressed as E180
stopifnot(
  !any(
    grepl(
      "-[NS][0-9]{2}E180_",
      fabdem_required_tiles$parent_folder
    )
  )
)


# Check the specific cases that caused problems earlier

fabdem_required_tiles[
  tile_name %chin%
    c(
      "N004W002",
      "S055W068"
    ),
  .(
    tile_name,
    parent_folder,
    mirror_file,
    url
  )
]


#/*==========================================================*/
#' ## 5. Directories
#/*==========================================================*/

fabdem_tmp_dir <- "Data/Raw/FABDEM/tmp"

fabdem_output_dir <- "Data/Processed/FABDEM/extracted_tiles"

dir.create(
  fabdem_tmp_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  fabdem_output_dir,
  recursive = TRUE,
  showWarnings = FALSE
)


fabdem_required_tiles[, 
  out_path := file.path(fabdem_output_dir, paste0( tile_name, ".rds"))
]

# A tile filename alone does not show whether it contains the current
# points. Old files can have the same name but different elev_point_id values.
tile_is_complete <- function(k) {
  path <- fabdem_required_tiles$out_path[k]
  if (!file.exists(path)) return(FALSE)

  saved <- tryCatch(as.data.table(readRDS(path)), error = function(e) NULL)
  if (is.null(saved)) return(FALSE)

  required <- fabdem_points_covered[
    tile_name == fabdem_required_tiles$tile_name[k]
  ]
  columns <- c(
    "elev_point_id", "point_type", "lon", "lat",
    "elevation_m", "elevation_dataset"
  )
  if (!all(columns %in% names(saved))) return(FALSE)
  if (nrow(saved) != nrow(required)) return(FALSE)

  setorder(saved, elev_point_id)
  setorder(required, elev_point_id)

  identical(saved$elev_point_id, required$elev_point_id) &&
    identical(as.character(saved$point_type),
              as.character(required$point_type)) &&
    identical(saved$lon, required$lon) &&
    identical(saved$lat, required$lat) &&
    isTRUE(all(saved$elevation_dataset == "FABDEM_V1-2"))
}


#/*==========================================================*/
#' ## 6. Remove stale temporary files from OLD runs
#/*==========================================================*/

# Run this only when no old FABDEM workers are running.

stale_tmp_files <-
  list.files(
    fabdem_tmp_dir,
    pattern = "\\.tif(\\.part)?$",
    full.names = TRUE
  )

if (length(stale_tmp_files) > 0L) {
  unlink(stale_tmp_files)
}


#/*==========================================================*/
#' ## 7. Function to process one FABDEM tile
#/*==========================================================*/

process_fabdem_tile <- function(
  k,
  max_download_attempts = 5L
) {

  tile <-
    fabdem_required_tiles[k]

  this_tile_name <-
    tile$tile_name

  out_path <-
    tile$out_path

  tif_path <-
    file.path(
      fabdem_tmp_dir,
      tile$mirror_file
    )

  part_path <-
    paste0(
      tif_path,
      ".part"
    )

  rds_part_path <-
    paste0(
      out_path,
      ".part"
    )


  #--------------------------------#
  # Skip completed tiles
  #--------------------------------#

  if (tile_is_complete(k)) {

    return(
      data.table(
        k = k,
        tile_name = this_tile_name,
        status = "already_done",
        error = NA_character_
      )
    )
  }


  #--------------------------------#
  # Select points in tile
  #--------------------------------#

  points_tile <-
    fabdem_points_covered[
      tile_name == this_tile_name
    ]

  if (nrow(points_tile) == 0L) {

    return(
      data.table(
        k = k,
        tile_name = this_tile_name,
        status = "no_points",
        error = NA_character_
      )
    )
  }


  #--------------------------------#
  # Guaranteed temporary cleanup
  #--------------------------------#

  on.exit(
    {

      if (file.exists(part_path)) {
        unlink(part_path)
      }

      if (file.exists(tif_path)) {
        unlink(tif_path)
      }

      if (file.exists(rds_part_path)) {
        unlink(rds_part_path)
      }

    },
    add = TRUE
  )


  # Remove remnants specific to this tile

  if (file.exists(part_path)) {
    unlink(part_path)
  }

  if (file.exists(tif_path)) {
    unlink(tif_path)
  }

  if (file.exists(rds_part_path)) {
    unlink(rds_part_path)
  }


  #--------------------------------#
  # Download with automatic retries
  #--------------------------------#

  download_success <-
    FALSE

  last_error <-
    NA_character_


  for (
    attempt in seq_len(max_download_attempts)
  ) {

    ok <-
      tryCatch(
        {

          download.file(
            tile$url,
            destfile = part_path,
            mode = "wb",
            method = "libcurl",
            quiet = TRUE
          )

          file.exists(part_path) &&
            !is.na(file.info(part_path)$size) &&
            file.info(part_path)$size > 0

        },
        error = function(e) {

          last_error <<-
            conditionMessage(e)

          FALSE
        }
      )


    if (isTRUE(ok)) {

      download_success <-
        TRUE

      break
    }


    if (file.exists(part_path)) {
      unlink(part_path)
    }


    # Backoff between failed attempts

    Sys.sleep(
      min(
        2^(attempt - 1),
        10
      ) +
        runif(
          1,
          0,
          1
        )
    )
  }


  if (!download_success) {

    return(
      data.table(
        k = k,
        tile_name = this_tile_name,
        status = "failed_download",
        error = last_error
      )
    )
  }


  #--------------------------------#
  # Promote completed download
  #--------------------------------#

  rename_ok <-
    file.rename(
      part_path,
      tif_path
    )

  if (!rename_ok) {

    return(
      data.table(
        k = k,
        tile_name = this_tile_name,
        status = "failed_file_rename",
        error = "Could not rename .tif.part to .tif"
      )
    )
  }


  #--------------------------------#
  # Open raster
  #--------------------------------#

  dem <-
    tryCatch(
      terra::rast(
        tif_path
      ),
      error = function(e) e
    )


  if (inherits(dem, "error")) {

    return(
      data.table(
        k = k,
        tile_name = this_tile_name,
        status = "failed_raster",
        error = conditionMessage(dem)
      )
    )
  }


  #--------------------------------#
  # Point vector
  #--------------------------------#

  points_vect <-
    tryCatch(
      terra::vect(
        points_tile,
        geom = c(
          "lon",
          "lat"
        ),
        crs = "EPSG:4326"
      ),
      error = function(e) e
    )


  if (inherits(points_vect, "error")) {

    return(
      data.table(
        k = k,
        tile_name = this_tile_name,
        status = "failed_vector",
        error = conditionMessage(points_vect)
      )
    )
  }


  #--------------------------------#
  # Extract elevation
  #--------------------------------#

  elev <-
    tryCatch(
      terra::extract(
        dem,
        points_vect,
        method = "bilinear"
      ),
      error = function(e) e
    )


  if (inherits(elev, "error")) {

    return(
      data.table(
        k = k,
        tile_name = this_tile_name,
        status = "failed_extract",
        error = conditionMessage(elev)
      )
    )
  }


  if (nrow(elev) != nrow(points_tile)) {

    return(
      data.table(
        k = k,
        tile_name = this_tile_name,
        status = "failed_extract",
        error = "Number of extracted values does not match number of points"
      )
    )
  }


  #--------------------------------#
  # Result
  #--------------------------------#

  result <-
    copy(
      points_tile
    )

  result[
    ,
    elevation_m :=
      elev[[2L]]
  ]

  result[
    ,
    elevation_dataset :=
      "FABDEM_V1-2"
  ]


  #--------------------------------#
  # Atomic RDS write
  #--------------------------------#

  saveRDS(
    result,
    rds_part_path,
    compress = FALSE
  )


  rds_rename_ok <-
    file.rename(
      rds_part_path,
      out_path
    )


  if (!rds_rename_ok) {

    return(
      data.table(
        k = k,
        tile_name = this_tile_name,
        status = "failed_rds_rename",
        error = "Could not rename temporary RDS output"
      )
    )
  }


  data.table(
    k = k,
    tile_name = this_tile_name,
    status = "success",
    error = NA_character_
  )
}


#/*==========================================================*/
#' ## 8. Run all unfinished tiles automatically
#/*==========================================================*/

# Catch errors for one tile without stopping all the other workers.
process_fabdem_tile_safe <- function(k) {
  tryCatch(
    process_fabdem_tile(k),
    error = function(e) {
      data.table(
        k = k,
        tile_name = fabdem_required_tiles$tile_name[k],
        status = "unexpected_error",
        error = conditionMessage(e)
      )
    }
  )
}

run_fabdem_extraction <- function(
  n_cores = 10L,
  max_rounds = 4L
) {

  # PSOCK workers start independent R sessions; they do not fork R on macOS.
  cl <- parallel::makePSOCKcluster(n_cores)
  on.exit(parallel::stopCluster(cl), add = TRUE)

  parallel::clusterEvalQ(cl, library(data.table))
  parallel::clusterExport(
    cl,
    varlist = c(
      "fabdem_required_tiles",
      "fabdem_points_covered",
      "fabdem_tmp_dir",
      "fabdem_output_dir",
      "tile_is_complete",
      "process_fabdem_tile",
      "process_fabdem_tile_safe"
    ),
    envir = .GlobalEnv
  )

  all_status <-
    vector(
      "list",
      max_rounds
    )


  for (
    round in seq_len(max_rounds)
  ) {

    completed <- vapply(
      seq_len(nrow(fabdem_required_tiles)),
      tile_is_complete,
      logical(1)
    )

    tiles_todo <-
      which(
        !completed
      )


    if (length(tiles_todo) == 0L) {

      message(
        "All FABDEM tiles completed."
      )

      break
    }


    message(
      sprintf(
        "Round %d: processing %d unfinished tiles",
        round,
        length(tiles_todo)
      )
    )


    round_status <- parallel::parLapplyLB(
      cl,
      tiles_todo,
      process_fabdem_tile_safe
    )


    round_status <-
      rbindlist(
        round_status,
        use.names = TRUE,
        fill = TRUE
      )


    round_status[
      ,
      round := round
    ]


    all_status[[round]] <-
      round_status


    print(
      round_status[
        ,
        .N,
        by = status
      ][
        order(-N)
      ]
    )


    # Brief pause before retry round

    if (round < max_rounds) {

      Sys.sleep(5)
    }
  }


  all_status <-
    rbindlist(
      all_status,
      use.names = TRUE,
      fill = TRUE
    )


  all_status
}


#/*==========================================================*/
#' ## 9. Execute
#/*==========================================================*/

# Four independent workers download and extract different tiles concurrently.
# Only files containing exactly the requested new points are reused.

fabdem_run_status <- if (nrow(fabdem_required_tiles)) {
  run_fabdem_extraction(
    n_cores = 10L,
    max_rounds = 4L
  )
} else {
  data.table(
    k = integer(),
    tile_name = character(),
    status = character(),
    error = character(),
    round = integer()
  )
}
  
# Save processing log

fwrite(
  fabdem_run_status,
  "Data/Processed/FABDEM/fabdem_run_status.csv"
)


#/*==========================================================*/
#' ## 10. FINAL tile-level validation
#/*==========================================================*/

fabdem_required_tiles[
  ,
  completed := vapply(
    seq_len(.N),
    tile_is_complete,
    logical(1)
  )
]


missing_tiles <-
  fabdem_required_tiles[
    completed == FALSE
  ]


cat(
  "Required tiles:",
  nrow(fabdem_required_tiles),
  "\n"
)

cat(
  "Completed tiles:",
  sum(fabdem_required_tiles$completed),
  "\n"
)

cat(
  "Missing tiles:",
  nrow(missing_tiles),
  "\n"
)


if (nrow(missing_tiles) > 0L) {

  fwrite(
    missing_tiles,
    "Data/Processed/FABDEM/fabdem_missing_tiles.csv"
  )

  stop(
    paste0(
      nrow(missing_tiles),
      " FABDEM tiles remain unresolved. ",
      "See fabdem_missing_tiles.csv."
    )
  )
}


#/*==========================================================*/
#' ## 11. Combine all tile-level elevation results
#/*==========================================================*/

fabdem_result_files <- fabdem_required_tiles$out_path


stopifnot(
  all(
    file.exists(
      fabdem_result_files
    )
  )
)


new_fabdem_elevation <- if (length(fabdem_result_files)) {
  rbindlist(lapply(fabdem_result_files, readRDS), use.names = TRUE)
} else {
  data.table(
    elev_point_id = integer(),
    point_type = character(),
    lon = numeric(),
    lat = numeric(),
    tile_name = character(),
    elevation_m = numeric(),
    elevation_dataset = character()
  )
}

reused_fabdem_elevation <- fabdem_points_lookup[
  fabdem_covered & is.finite(cached_elevation_m) &
    cached_elevation_dataset == "FABDEM_V1-2",
  .(
    elev_point_id,
    point_type,
    lon,
    lat,
    elevation_m = cached_elevation_m,
    elevation_dataset = cached_elevation_dataset
  )
]

fabdem_elevation <- rbindlist(
  list(new_fabdem_elevation, reused_fabdem_elevation),
  use.names = TRUE,
  fill = TRUE
)


setorder(
  fabdem_elevation,
  elev_point_id
)


#/*==========================================================*/
#' ## 12. FINAL point-level validation
#/*==========================================================*/

# No duplicate point IDs

duplicate_ids <-
  fabdem_elevation[,
    .N,
    by = elev_point_id
  ][
    N > 1L
  ]

stopifnot(
  nrow(duplicate_ids) == 0L
)


# Every newly extracted FABDEM point must have an output row.
missing_point_ids <-
  setdiff(
    fabdem_points_covered$elev_point_id,
    new_fabdem_elevation$elev_point_id
  )

stopifnot(
  length(missing_point_ids) == 0L
)


cat(
  "New FABDEM points:",
  nrow(fabdem_points_covered),
  "\n"
)

cat(
  "Extracted point records:",
  nrow(fabdem_elevation),
  "\n"
)

cat(
  "Unique extracted point IDs:",
  uniqueN(
    fabdem_elevation$elev_point_id
  ),
  "\n"
)

cat(
  "NA elevation values:",
  sum(
    is.na(
      fabdem_elevation$elevation_m
    )
  ),
  "\n"
)


#/*==========================================================*/
#' ## 13. Save completed FABDEM elevation table
#/*==========================================================*/
saveRDS(
  fabdem_elevation,
  "Data/Processed/FABDEM/fabdem_elevation.rds",
  compress = FALSE
)
```


# Work on Copernicus GLO-30

Next, build a complete elevation lookup, fill only the remaining gaps with Copernicus GLO-30, then attach elevations back to uc_2025 and surface_water_candidates.

```{r}
# /*===========================================*/
#'=  Build the master elevation table =
# /*===========================================*/
fabdem_elevation <- readRDS("Data/Processed/FABDEM/fabdem_elevation.rds")

elevation_points <- copy(
  fabdem_points_lookup[, .(
    elev_point_id,
    point_type,
    lon,
    lat,
    fabdem_covered,
    elevation_m = cached_elevation_m,
    elevation_dataset = cached_elevation_dataset
  )]
)

elevation_points[
  fabdem_elevation[, .(
    elev_point_id,
    elevation_m,
    elevation_dataset
  )],
  on = "elev_point_id",
  `:=`(
    elevation_m = i.elevation_m,
    elevation_dataset = i.elevation_dataset
  )
]

```


```{r}
#/*--------------------------------*/
#' ## Check what remains missing
#/*--------------------------------*/
elevation_points[, .N, by = .(
  point_type,
  fabdem_covered,
  missing_elevation = is.na(elevation_m)
)]

copdem_points <- elevation_points[
  is.na(elevation_m)
]

nrow(copdem_points)

copdem_points[, .N, by = .(
  point_type,
  fabdem_covered
)]
```



```{r}
# /*===========================================*/
#'=  Extract Copernicus GLO-30 elevations =
# /*===========================================*/
copdem_points <- elevation_points[!is.finite(elevation_m)]

if (nrow(copdem_points) > 0L) {
library(rgee)

# I did
# conda activate tools
# conda install -c conda-forge earthengine-api numpy

# Then, restart R session completely because because reticulate cannot reliably switch Python environments after Python has already been initialized.
# In the new R session, run:
# library(reticulate)

# use_python(
#   "/Users/shunkeikakimoto/miniforge3/envs/tools/bin/python3.11",
#   required = TRUE
# )

# py_config()
# py_module_available("ee")

# ee_install_set_pyenv(
#   py_path = "/Users/shunkeikakimoto/miniforge3/envs/tools/bin/python3.11",
#   py_env = "tools"
# )



ee_Authenticate(
  auth_mode = "notebook"
)

ee$Initialize(
  project = "gep-water-2026"
)

# Extract only points still missing after reuse and FABDEM.

nrow(copdem_points)

#/*--------------------------------*/
#' ## load the current Copernicus GLO-30 2024_1 DEM
#/*--------------------------------*/
copdem_collection <- ee$ImageCollection(
  "COPERNICUS/DEM/GLO30_2024_1"
)

native_projection <- copdem_collection$
  first()$
  select("DEM")$
  projection()

copdem <- copdem_collection$
  mosaic()$
  setDefaultProjection(native_projection)$
  select("DEM")$
  rename("elevation_m")


#/*--------------------------------*/
#' ## Function to extract one batch
#/*--------------------------------*/
extract_copdem_batch <- function(dt) {

  features <- lapply(seq_len(nrow(dt)), function(j) {

    ee$Feature(
      ee$Geometry$Point(
        list(
          dt$lon[j],
          dt$lat[j]
        )
      ),
      list(
        elev_point_id = as.character(dt$elev_point_id[j])
      )
    )
  })

  points_fc <- ee$FeatureCollection(
    features
  )

  reducer <- ee$Reducer$
    first()$
    setOutputs(
      list("elevation_m")
    )

  result <- copdem$
    reduceRegions(
      collection = points_fc,
      reducer = reducer,
      scale = 30
    )$
    getInfo()

  rbindlist(
    lapply(result$features, function(x) {

      p <- x$properties

      data.table(
        elev_point_id = as.integer(p$elev_point_id),
        elevation_m = if (is.null(p$elevation_m)) {
          NA_real_
        } else {
          as.numeric(p$elevation_m)
        }
      )
    })
  )
}


#/*--------------------------------*/
#' ## Extract all fallback points
#/*--------------------------------*/
batch_size <- 200L

copdem_points[, batch_id :=
  ceiling(.I / batch_size)]

copdem_results <- lapply(
  split(
    copdem_points,
    by = "batch_id",
    keep.by = FALSE
  ),
  extract_copdem_batch
)

copdem_elevation <- 
  rbindlist(
    copdem_results,
    use.names = TRUE
  )

setorder(
  copdem_elevation,
  elev_point_id
)


# Identify the dataset
copdem_elevation[, elevation_dataset := "COPERNICUS_GLO30_2024_1"]

# Validate:
nrow(copdem_elevation)

uniqueN(copdem_elevation$elev_point_id)

copdem_elevation[is.na(elevation_m), .N ]

summary(copdem_elevation$elevation_m)

#/*--------------------------------*/
#' ## Fill the missing elevations
#/*--------------------------------*/
elevation_points[
  copdem_elevation,
  on = "elev_point_id",
  `:=`(
    elevation_m = i.elevation_m,
    elevation_dataset = i.elevation_dataset
  )
]

} # No Earth Engine session is needed if all elevations were reused.

# Check
remaining_missing <- elevation_points[!is.finite(elevation_m), .N]
if (remaining_missing > 0L) {
  stop(remaining_missing, " requested points still lack elevation")
}

stopifnot(
  !anyDuplicated(elevation_points[, .(point_type, lon, lat)]),
  nrow(elevation_points) == nrow(fabdem_points_lookup)
)


# === Save === #
saveRDS(
  elevation_points,
  "Data/Processed/FABDEM/elevation_points_complete.rds",
  compress = FALSE
)
```
