---
title: "Country totals for municipal raw-water replacement cost"
format: html
execute:
  echo: true
---

This script aggregates the **shared-flow-path** replacement results from step 8 by country. It uses the city-to-country identifiers in the prepared groundwater input file. Country cost totals cover only cities with feasible baseline and replacement portfolios. The output retains counts of cities without a value.  No cost is assigned to those cities.

## 1. Read the city results and country identifiers

```{r}

analysis_dir <- here("Data/Analysis")
result_dir <- here("Data/Results")

if(!dir.exists(result_dir)) {
  dir.create(result_dir, recursive = TRUE)
}

scenario_tag <- "30"

# === City-level replacement cost results === #
city_result <- fread(
    file.path(
      analysis_dir,
      paste0("y_city_shared_flowpath_replacement_", scenario_tag, ".csv")
    )
  )


# === City-to-Country lookup === #
uc_2025 <- readRDS(
    here("Data/Processed/ghsl_urban_centres_2025.rds")
  )

city_iso3_lookup <- 
  uc_2025[
    , iso3 := countrycode::countrycode(country_id, "iso3n", "iso3c")
  ][
    , .(
      iso3,
      urban_id
    )
  ]

stopifnot(!anyDuplicated(city_iso3_lookup$urban_id))

# === Country name template === #
country_template <- 
  st_read(
    here("Data/Raw/Boundary/z_ee_r250_correspondence.gpkg"),
    quiet = TRUE
  ) %>%
  st_drop_geometry() %>%
  as.data.table() %>%
  .[
    , .(
      iso3_r250_id, 
      iso3_r250_label,
      iso3_r250_name
      )
  ]

stopifnot(!anyDuplicated(country_template$iso3_r250_id))
```

# 2. Attach country identifiers to city-level results
```{r}

required_city_columns <- c(
  "urban_id",
  "replacement_status",
  "demand_m3_year",
  "baseline_cost_2019usd_year",
  "replacement_cost_2019usd_year",
  "extra_cost_2019usd_year",
  "has_valid_city_electricity_price"
)


city_result <- 
  merge(
    city_result[, ..required_city_columns],
    city_iso3_lookup,
    by = "urban_id",
    all.x = TRUE,
    sort = FALSE
  )

stopifnot(
  nrow(city_result) == uniqueN(city_result$urban_id),
  !anyNA(city_result$iso3),
  !anyNA(city_result$country_name),
  !anyNA(city_result$country_id),
  all(nchar(city_result$iso3) == 3L),
  all(is.finite(city_result$demand_m3_year) & city_result$demand_m3_year > 0)
)

city_result[,
  valid_electricity_price :=
    as.logical(has_valid_city_electricity_price)
]

stopifnot(!anyNA(city_result$valid_electricity_price))
```


## 3. Aggregate valued cities and report coverage

```{r}
# Identify cities with a feasible baseline and replacement.
city_result[
  , valued := replacement_status == "feasible_replacement"
]

stopifnot(
  !anyNA(city_result$iso3),
  !anyNA(city_result$valued),
  !anyNA(city_result$valid_electricity_price),
  all(city_result$replacement_status %in% c(
    "feasible_replacement",
    "baseline_uncovered",
    "no_feasible_replacement"
  )),
  all(is.finite(
    city_result[valued == TRUE]$baseline_cost_2019usd_year
  )),
  all(is.finite(
    city_result[valued == TRUE]$replacement_cost_2019usd_year
  )),
  all(is.finite(
    city_result[valued == TRUE]$extra_cost_2019usd_year
  ))
)

# Aggregate city results by ISO3.
country_results <- city_result[
  , .(
    n_cities = .N,
    n_valued_cities = sum(valued),

    n_baseline_uncovered_missing_price = sum(
      replacement_status == "baseline_uncovered" &
        !valid_electricity_price
    ),

    n_baseline_uncovered_valid_price = sum(
      replacement_status == "baseline_uncovered" &
        valid_electricity_price
    ),

    n_no_feasible_replacement = sum(
      replacement_status == "no_feasible_replacement"
    ),

    demand_m3_year_all_cities = sum(demand_m3_year),

    demand_m3_year_valued_cities = sum(
      demand_m3_year[valued]
    ),

    baseline_cost_2019usd_year = if (any(valued)) {
      sum(baseline_cost_2019usd_year[valued])
    } else {
      NA_real_
    },

    replacement_cost_2019usd_year = if (any(valued)) {
      sum(replacement_cost_2019usd_year[valued])
    } else {
      NA_real_
    },

    replacement_value_2019usd_year = if (any(valued)) {
      sum(extra_cost_2019usd_year[valued])
    } else {
      NA_real_
    }
  ),
  by = iso3
]

country_results[
  , `:=`(
    n_cities_without_value = n_cities - n_valued_cities,

    share_cities_valued = n_valued_cities / n_cities,

    value_coverage = fifelse(
      n_valued_cities == n_cities,
      "all_cities_valued",
      fifelse(
        n_valued_cities == 0L,
        "no_cities_valued",
        "partial"
      )
    )
  )
]

# Check that the template has one row per ISO3 and contains
# every ISO3 represented in the city results.
stopifnot(
  !anyDuplicated(country_template$iso3_r250_label)
)

unmatched_iso3 <- setdiff(
  country_results$iso3,
  country_template$iso3_r250_label
)

if (length(unmatched_iso3) > 0L) {
  stop(
    "Country results absent from the template: ",
    paste(unmatched_iso3, collapse = ", ")
  )
}

# Retain every template country, including those with no
# city in the analyzed sample.
country_results_final <- merge(
  country_template,
  country_results,
  by.x = "iso3_r250_label",
  by.y = "iso3",
  all.x = TRUE,
  sort = FALSE
)

stopifnot(
  nrow(country_results_final) == nrow(country_template)
)

# Explain why a country has no reported replacement value.
country_results_final[
  , missing_value_reason := fcase(
    is.na(n_cities),
    "no_city_in_analyzed_sample",

    n_valued_cities > 0,
    NA_character_,

    n_baseline_uncovered_missing_price == n_cities,
    "all_cities_baseline_uncovered_missing_electricity_price",

    n_baseline_uncovered_valid_price == n_cities,
    "all_cities_baseline_uncovered_valid_electricity_price",

    n_no_feasible_replacement == n_cities,
    "all_cities_no_feasible_replacement",

    default = "multiple_reasons_no_valued_cities"
  )
]

setcolorder(
  country_results_final,
  c(
    "iso3_r250_id",
    "iso3_r250_label",
    "iso3_r250_name",
    "n_cities",
    "n_valued_cities",
    "n_cities_without_value",
    "share_cities_valued",
    "value_coverage",
    "missing_value_reason",
    "n_baseline_uncovered_missing_price",
    "n_baseline_uncovered_valid_price",
    "n_no_feasible_replacement",
    "demand_m3_year_all_cities",
    "demand_m3_year_valued_cities",
    "baseline_cost_2019usd_year",
    "replacement_cost_2019usd_year",
    "replacement_value_2019usd_year"
  )
)

setorder(
  country_results_final,
  iso3_r250_id
)

# Validate totals using only countries with analyzed cities.
countries_with_cities <- 
  country_results_final[!is.na(n_cities) ]


stopifnot(
  sum(countries_with_cities$n_cities) ==
    nrow(city_result),

  sum(countries_with_cities$n_valued_cities) ==
    sum(city_result$valued),

  all(
    countries_with_cities$n_cities_without_value ==
      countries_with_cities$n_baseline_uncovered_missing_price +
      countries_with_cities$n_baseline_uncovered_valid_price +
      countries_with_cities$n_no_feasible_replacement
  ),

  isTRUE(all.equal(
    sum(
      countries_with_cities$replacement_value_2019usd_year,
      na.rm = TRUE
    ),
    sum(
      city_result[
        valued == TRUE,
        extra_cost_2019usd_year
      ]
    ),
    tolerance = 1e-10
  ))
)
```

## 4. Save the country results

```{r}
output_file <- file.path(
  result_dir,
  paste0("y_country_municipal_replacement_cost_", scenario_tag, ".csv")
)

fwrite(country_results_final, output_file)

# Number of countries by coverage and missing-value reason
print(
  country_results_final[
    , .N,
    by = .(value_coverage, missing_value_reason)
  ]
)

# Overall totals; the monetary total covers valued cities only
print(
  country_results_final[
    , .(
      template_countries = .N,
      countries_with_analyzed_cities = sum(!is.na(n_cities)),
      countries_with_reported_value = sum(
        !is.na(replacement_value_2019usd_year)
      ),
      analyzed_cities = sum(n_cities, na.rm = TRUE),
      valued_cities = sum(n_valued_cities, na.rm = TRUE),
      replacement_value_2019usd_year = sum(
        replacement_value_2019usd_year,
        na.rm = TRUE
      )
    )
  ]
)

message("Saved ", output_file)
```


## 5. For final report table

We only need iso3_r250_id, iso3_r250_label, iso3_r250_name, and wateruse_municipal_gep.

```{r}
final_gep_report <- 
  country_results_final[
    , .(
      iso3_r250_id,
      iso3_r250_label,
      iso3_r250_name,
      wateruse_municipal_gep = replacement_value_2019usd_year 
    )
  ]

fwrite(
  final_gep_report,
  file.path(
    result_dir,
    "wateruse_municipal_gep.csv"
  )
)

sum(final_gep_report$wateruse_municipal_gep, na.rm = TRUE)
```