---
title: "Cost of replacing each city's selected water sources"
format: html
execute:
  echo: true
---

This analysis starts with the **covered** city-level solutions saved by `7_city_water_cost_min.qmd`. For each city, it removes every surface-water source selected in step 8. If the city used groundwater, it also removes its groundwater option. It then finds the lowest-cost way to meet the *same* annual demand from the **remaining priced sources**.

Here a "source" means a shortlisted river reach or lake candidate ID.  Another intake on the same river may remain available. The cost difference therefore concerns the selected *candidate connections*, not necessarily loss of access to an entire river system.

The difference between the two annual costs measures the modeled cost of losing the city's selected source portfolio. Replacement water still comes from natural sources. This difference is not the value of all natural water provisioning and should not be added to the shared-flow-path replacement estimate from step 8. These are alternative counterfactuals.  An infeasible replacement receives `NA`, not zero or infinity. The final city output adds `missing_value_reason` and input-coverage fields. These labels distinguish missing city electricity data from a shortage among priced candidate sources; they do not change the model's cost.

## 1. Read the current 30% scenario and the saved step 7 solution

```{r}
library(data.table)
library(here)

if (!requireNamespace("highs", quietly = TRUE)) {
  stop("Install the highs R package before running this file")
}

analysis_dir <- here("Data/Analysis")
scenario_tag <- "30"
gw_scenario <- "0.3PD_0.25DL_0.2RR"
seconds_per_year <- 365.25 * 24 * 60 * 60

option_file <- file.path(
  analysis_dir, "city_surface_water_capacity_options.csv"
)

conflict_file <- file.path(
  analysis_dir, "surface_water_conflicts.rds"
)

groundwater_file <- file.path(
  analysis_dir, "ghsl_cities_gw_cost_compressed_20260921.csv"
)

price_file <- here(
  "Data/Processed/uc_2025_with_electricity_price.rds"
)

city_file <- file.path(
  analysis_dir,
  paste0("x_city_municipal_water_supply_", scenario_tag, ".csv")
)

chosen_file <- file.path(
  analysis_dir,
  paste0("x_city_surface_water_chosen_designs_", scenario_tag, ".csv")
)

input_files <- c(
  option_file, conflict_file, groundwater_file, city_file, chosen_file,
  price_file
)

if (!all(file.exists(input_files))) {
  stop(
    "Missing input file(s): ",
    paste(input_files[!file.exists(input_files)], collapse = ", ")
  )
}

options <- fread(option_file)
conflicts <- as.data.table(readRDS(conflict_file))
city <- fread(city_file)
chosen <- fread(chosen_file)

gw_cost_col <- paste0("gw_unit_cost_2019_usd_m3_", gw_scenario)
gw_capacity_col <- paste0("gw_capacity_m3_year_", gw_scenario)

groundwater <- fread(
  groundwater_file,
  select = c("urban_id", gw_cost_col, gw_capacity_col)
)

setnames(
  groundwater,
  c(gw_cost_col, gw_capacity_col),
  c("gw_cost_2019usd_m3", "gw_capacity_m3_year")
)

required_options <- c(
  "urban_id", "source_key", "design_id", "source_type",
  "source_id", "max_delivery_m3_year",
  "source_capacity_m3_year", "fixed_cost_2019usd_year",
  "variable_cost_2019usd_m3", "withdrawal_fraction",
  "capacity_flow_basis", "source_discharge_m3s"
)
required_city <- c(
  "urban_id", "status", "withdrawal_fraction",
  "demand_m3_year", "surface_m3_year", "groundwater_m3_year",
  "total_cost_2019usd_year"
)
required_chosen <- c(
  "urban_id", "source_key", "design_id", "quantity_m3_year"
)

stopifnot(
  all(required_options %in% names(options)),
  all(required_city %in% names(city)),
  all(required_chosen %in% names(chosen)),
  all(c("urban_id", "source_key_a", "source_key_b") %in%
        names(conflicts)),
  !anyDuplicated(city$urban_id),
  !anyDuplicated(groundwater$urban_id),
  !anyDuplicated(options[, .(urban_id, source_key, design_id)]),
  !anyDuplicated(chosen[, .(urban_id, source_key)]),
  all(city$status %in% c("covered", "uncovered")),
  all(is.finite(city$demand_m3_year) & city$demand_m3_year > 0),
  uniqueN(options$withdrawal_fraction) == 1L,
  abs(unique(options$withdrawal_fraction) - 0.30) < 1e-8,
  all(abs(city$withdrawal_fraction - 0.30) < 1e-8),
  uniqueN(options$capacity_flow_basis) == 1L,
  unique(options$capacity_flow_basis) ==
    "mean_annual_naturalized_discharge",
  all(is.finite(options$max_delivery_m3_year)),
  all(options$max_delivery_m3_year > 0),
  all(is.finite(options$fixed_cost_2019usd_year)),
  all(options$fixed_cost_2019usd_year >= 0),
  all(is.finite(options$variable_cost_2019usd_m3)),
  all(options$variable_cost_2019usd_m3 >= 0),
  all(abs(
    options$source_capacity_m3_year -
      0.30 * options$source_discharge_m3s * seconds_per_year
  ) <= pmax(1e-2, 1e-7 * options$source_capacity_m3_year)),
  all(chosen$urban_id %in% city[status == "covered", urban_id])
)

setkey(options, urban_id)
setkey(conflicts, urban_id)
setkey(chosen, urban_id)
setkey(groundwater, urban_id)

solver_control <- highs::highs_control(
  threads = 1L,
  log_to_console = FALSE,
  mip_rel_gap = 1e-8,
  mip_abs_gap = 0,
  presolve = "off"
)

# Retry only if the second minimum appears cheaper than the saved optimum.
retry_control <- highs::highs_control(
  threads = 1L,
  log_to_console = FALSE,
  mip_rel_gap = 1e-8,
  mip_abs_gap = 0,
  presolve = "on"
)
```

## 2. Build the same supply problem as step 7

Continuous variables represent delivered water as shares of city demand.  Binary variables choose at most one pipe design per source. A connection cannot deliver more than its design permits. Network conflicts and the groundwater capacity are retained. Only the available source set changes.

```{r}
make_city_model <- function(
  city_options,
  city_conflicts,
  demand,
  gw_cost,
  gw_capacity
) {
  n_options <- nrow(city_options)
  sources <- unique(city_options$source_key)
  n_sources <- length(sources)

  flow <- seq_len(n_options)
  built <- n_options + seq_len(n_options)
  groundwater <- 2L * n_options + 1L

  n_rows <- 1L + n_options + n_sources +
    nrow(city_conflicts) + 1L

  A <- matrix(0, nrow = n_rows, ncol = groundwater)
  rhs <- numeric(n_rows)

  A[1L, c(flow, groundwater)] <- 1
  rhs[1L] <- 1

  for (j in seq_len(n_options)) {
    row <- 1L + j
    A[row, flow[j]] <- 1
    A[row, built[j]] <- -min(
      1,
      city_options$max_delivery_m3_year[j] / demand
    )
  }

  for (j in seq_along(sources)) {
    row <- 1L + n_options + j
    A[row, built[city_options$source_key == sources[j]]] <- 1
    rhs[row] <- 1
  }

  for (j in seq_len(nrow(city_conflicts))) {
    row <- 1L + n_options + n_sources + j
    pair <- c(
      city_conflicts$source_key_a[j],
      city_conflicts$source_key_b[j]
    )
    A[row, built[city_options$source_key %in% pair]] <- 1
    rhs[row] <- 1
  }

  A[n_rows, groundwater] <- 1
  rhs[n_rows] <- gw_capacity / demand

  objective <- c(
    city_options$variable_cost_2019usd_m3,
    city_options$fixed_cost_2019usd_year / demand,
    if (gw_capacity > 0) gw_cost else 0
  )

  types <- rep("C", length(objective))
  types[built] <- "I"

  list(
    A = A,
    rhs = rhs,
    objective = objective,
    types = types,
    flow = flow,
    built = built,
    groundwater = groundwater,
    demand = demand,
    options = city_options
  )
}

solve_city_model <- function(
  model,
  require_full_demand = TRUE,
  control = solver_control
) {
  if (require_full_demand) {
    objective <- model$objective
  } else {
    # Find the maximum delivered water when full replacement is infeasible.
    objective <- numeric(length(model$objective))
    objective[c(model$flow, model$groundwater)] <- 1
  }

  lhs <- rep(-Inf, nrow(model$A))
  if (require_full_demand) {
    lhs[1L] <- 1
  }

  fit <- highs::highs_solve(
    L = objective,
    lower = rep(0, length(objective)),
    upper = rep(1, length(objective)),
    A = model$A,
    lhs = lhs,
    rhs = model$rhs,
    types = model$types,
    maximum = !require_full_demand,
    control = control
  )

  if (identical(fit$status_message, "Infeasible")) {
    return(list(status = "infeasible"))
  }
  if (!identical(fit$status_message, "Optimal")) {
    stop("Solver status: ", fit$status_message)
  }

  solution <- fit$primal_solution
  residual <- as.vector(model$A %*% solution - model$rhs)
  tolerance <- 1e-6

  if (
    (require_full_demand && abs(residual[1L]) > tolerance) ||
    (!require_full_demand && residual[1L] > tolerance) ||
    any(residual[-1L] > tolerance) ||
    any(abs(solution[model$built] -
              round(solution[model$built])) > tolerance)
  ) {
    stop("Solver returned a solution that fails model constraints")
  }

  if (require_full_demand) {
    list(
      status = "optimal",
      cost = fit$objective_value * model$demand,
      solution = solution
    )
  } else {
    list(
      status = "optimal",
      max_supply = min(
        model$demand,
        pmax(0, fit$objective_value * model$demand)
      )
    )
  }
}
```

## 3. Remove the selected sources and find the cheapest replacement

The saved step 7 portfolio is the baseline. Before excluding its sources, the script rebuilds its annual cost from the current pipe options and groundwater data. A disagreement stops the run rather than comparing different versions of the model.

```{r}
city_parts <- vector("list", nrow(city))
source_parts <- vector("list", nrow(city))

for (i in seq_len(nrow(city))) {
  baseline <- city[i]
  id <- baseline$urban_id
  demand <- baseline$demand_m3_year

  if (baseline$status != "covered") {
    city_parts[[i]] <- data.table(
      urban_id = id,
      withdrawal_fraction = 0.30,
      replacement_status = "baseline_uncovered",
      demand_m3_year = demand,
      n_removed_surface_sources = NA_integer_,
      removed_groundwater = NA,
      n_remaining_surface_sources = NA_integer_,
      baseline_cost_2019usd_year = NA_real_,
      replacement_max_supply_m3_year = NA_real_,
      replacement_shortage_m3_year = NA_real_,
      replacement_cost_2019usd_year = NA_real_,
      extra_cost_2019usd_year = NA_real_,
      replacement_surface_m3_year = NA_real_,
      replacement_groundwater_m3_year = NA_real_,
      n_replacement_surface_connections = NA_integer_,
      solver_retry_presolve_on = FALSE
    )
    next
  }

  city_options <- options[.(id), nomatch = 0L]
  selected <- chosen[.(id), nomatch = 0L]
  city_gw <- groundwater[.(id), nomatch = 0L]

  gw_cost <- if (nrow(city_gw)) city_gw$gw_cost_2019usd_m3 else NA_real_
  gw_capacity <- if (nrow(city_gw)) city_gw$gw_capacity_m3_year else NA_real_

  if (
    !is.finite(gw_cost) || gw_cost < 0 ||
    !is.finite(gw_capacity) || gw_capacity <= 0
  ) {
    gw_cost <- 0
    gw_capacity <- 0
  }
  gw_capacity <- min(gw_capacity, demand)

  if (
    !is.finite(baseline$total_cost_2019usd_year) ||
    baseline$total_cost_2019usd_year < 0 ||
    !is.finite(baseline$groundwater_m3_year) ||
    baseline$groundwater_m3_year < -1e-6 ||
    baseline$groundwater_m3_year >
      gw_capacity + pmax(1e-3, 1e-7 * demand)
  ) {
    stop("Invalid saved step 7 solution at urban_id ", id)
  }

  matched <- city_options[
    selected[, .(urban_id, source_key, design_id)],
    on = .(urban_id, source_key, design_id)
  ]

  if (
    nrow(matched) != nrow(selected) ||
    any(!is.finite(matched$max_delivery_m3_year)) ||
    any(!is.finite(selected$quantity_m3_year)) ||
    any(selected$quantity_m3_year < -1e-6) ||
    any(selected$quantity_m3_year >
          matched$max_delivery_m3_year +
            pmax(1e-3, 1e-7 * demand))
  ) {
    stop("Saved designs do not match the options at urban_id ", id)
  }

  surface_quantity <- sum(selected$quantity_m3_year)
  reconstructed_cost <- sum(
    matched$fixed_cost_2019usd_year +
      matched$variable_cost_2019usd_m3 *
        selected$quantity_m3_year
  ) + gw_cost * baseline$groundwater_m3_year

  if (
    abs(surface_quantity - baseline$surface_m3_year) >
      pmax(1e-3, 1e-6 * demand) ||
    abs(surface_quantity + baseline$groundwater_m3_year - demand) >
      pmax(1e-3, 1e-6 * demand) ||
    abs(reconstructed_cost - baseline$total_cost_2019usd_year) >
      pmax(0.05, 1e-6 * baseline$total_cost_2019usd_year)
  ) {
    stop("Step 7 results disagree with current inputs at urban_id ", id)
  }

  # Exclude the sources themselves, not merely the chosen pipe sizes.
  removed_keys <- unique(selected$source_key)
  removed_groundwater <- baseline$groundwater_m3_year > 1e-6

  remaining <- city_options[!source_key %in% removed_keys]
  remaining_keys <- unique(remaining$source_key)

  city_conflicts <- conflicts[.(id), nomatch = 0L]
  city_conflicts <- city_conflicts[
    source_key_a %in% remaining_keys &
      source_key_b %in% remaining_keys
  ]

  replacement_gw_capacity <- if (removed_groundwater) 0 else gw_capacity

  model <- make_city_model(
    city_options = remaining,
    city_conflicts = city_conflicts,
    demand = demand,
    gw_cost = gw_cost,
    gw_capacity = replacement_gw_capacity
  )

  result <- tryCatch(
    solve_city_model(model),
    error = function(e) {
      stop("Replacement solve at urban_id ", id, ": ", conditionMessage(e))
    }
  )

  replacement_cost <- NA_real_
  extra_cost <- NA_real_
  replacement_surface <- NA_real_
  replacement_groundwater <- NA_real_
  n_replacement_connections <- NA_integer_
  used_retry <- FALSE

  if (result$status == "optimal") {
    cost_tolerance <- pmax(
      0.05,
      1e-6 * baseline$total_cost_2019usd_year
    )

    if (result$cost < baseline$total_cost_2019usd_year - cost_tolerance) {
      # A restricted choice set cannot have a cheaper optimum. Check both
      # problems with the alternate solver control before accepting a cost.
      all_keys <- unique(city_options$source_key)
      baseline_conflicts <- conflicts[.(id), nomatch = 0L][
        source_key_a %in% all_keys &
          source_key_b %in% all_keys
      ]
      baseline_model <- make_city_model(
        city_options = city_options,
        city_conflicts = baseline_conflicts,
        demand = demand,
        gw_cost = gw_cost,
        gw_capacity = gw_capacity
      )

      original_retry <- solve_city_model(
        baseline_model,
        control = retry_control
      )
      replacement_retry <- solve_city_model(
        model,
        control = retry_control
      )

      if (
        original_retry$status != "optimal" ||
        replacement_retry$status != "optimal" ||
        abs(original_retry$cost -
              baseline$total_cost_2019usd_year) > cost_tolerance ||
        replacement_retry$cost <
          baseline$total_cost_2019usd_year - cost_tolerance
      ) {
        stop(
          "Restricted portfolio is cheaper than the saved step 7 ",
          "optimum at urban_id ", id,
          "; verify solver convergence and input versions"
        )
      }

      result <- replacement_retry
      used_retry <- TRUE
    }

    selected_rows <- which(result$solution[model$built] > 0.5)
    delivered <- result$solution[model$flow][selected_rows] * demand

    replacement_surface <- sum(delivered)
    replacement_groundwater <-
      result$solution[model$groundwater] * demand
    replacement_cost <- result$cost
    extra_cost <- pmax(
      0,
      replacement_cost - baseline$total_cost_2019usd_year
    )
    n_replacement_connections <- length(selected_rows)
    replacement_max_supply <- demand
    replacement_shortage <- 0
    status <- "feasible_replacement"

    if (length(selected_rows)) {
      source_parts[[i]] <- copy(remaining[selected_rows, .(
        urban_id, source_key, source_type, source_id, design_id,
        max_delivery_m3_year, fixed_cost_2019usd_year,
        variable_cost_2019usd_m3
      )])

      source_parts[[i]][, `:=`(
        quantity_m3_year = delivered,
        source_cost_2019usd_year =
          fixed_cost_2019usd_year +
            variable_cost_2019usd_m3 * delivered
      )]
    }
  } else {
    maximum <- solve_city_model(
      model,
      require_full_demand = FALSE
    )
    if (maximum$status != "optimal") {
      stop("Could not determine replacement supply at urban_id ", id)
    }

    replacement_max_supply <- maximum$max_supply
    replacement_shortage <- pmax(0, demand - replacement_max_supply)
    status <- "no_feasible_replacement"
  }

  city_parts[[i]] <- data.table(
    urban_id = id,
    withdrawal_fraction = 0.30,
    replacement_status = status,
    demand_m3_year = demand,
    n_removed_surface_sources = length(removed_keys),
    removed_groundwater = removed_groundwater,
    n_remaining_surface_sources = length(remaining_keys),
    baseline_cost_2019usd_year = baseline$total_cost_2019usd_year,
    replacement_max_supply_m3_year = replacement_max_supply,
    replacement_shortage_m3_year = replacement_shortage,
    replacement_cost_2019usd_year = replacement_cost,
    extra_cost_2019usd_year = extra_cost,
    replacement_surface_m3_year = replacement_surface,
    replacement_groundwater_m3_year = replacement_groundwater,
    n_replacement_surface_connections = n_replacement_connections,
    solver_retry_presolve_on = used_retry
  )

  if (i %% 250L == 0L) {
    message("Compared ", i, " / ", nrow(city), " cities")
  }
}
```

## 4. Check and save the results

`extra_cost_2019usd_year` is defined only where the remaining sources meet the entire city demand. The second file identifies which sources provide that replacement. Neither file treats a shortage as a monetary value.

```{r}
city_results <- rbindlist(city_parts, use.names = TRUE)

# Retain the modeled feasibility status and report missing input data
# separately. These fields do not change the optimization or the costs.
prices <- as.data.table(readRDS(price_file))[
  , .(urban_id, electricity_price_2019usd_kwh)
]
stopifnot(!anyDuplicated(prices$urban_id))

city_results[
  prices,
  electricity_price_2019usd_kwh :=
    i.electricity_price_2019usd_kwh,
  on = "urban_id"
]

city_results[
  groundwater,
  `:=`(
    gw_cost_input_2019usd_m3 = i.gw_cost_2019usd_m3,
    gw_capacity_input_m3_year = i.gw_capacity_m3_year
  ),
  on = "urban_id"
]

city_results[, has_valid_city_electricity_price :=
  is.finite(electricity_price_2019usd_kwh) &
    electricity_price_2019usd_kwh >= 0
]

city_results[, groundwater_input_status := fcase(
  !is.finite(gw_capacity_input_m3_year),
  "missing_groundwater_capacity",
  gw_capacity_input_m3_year <= 0,
  "no_positive_groundwater_capacity",
  !is.finite(gw_cost_input_2019usd_m3) |
    gw_cost_input_2019usd_m3 < 0,
  "missing_groundwater_cost",
  default = "usable_groundwater_cost_and_capacity"
)]

city_results[, missing_value_reason := fcase(
  replacement_status == "baseline_uncovered" &
    !has_valid_city_electricity_price,
  "baseline_uncovered_with_missing_city_electricity_price",
  replacement_status == "baseline_uncovered",
  "baseline_uncovered_with_valid_city_electricity_price",
  replacement_status == "no_feasible_replacement" &
    !has_valid_city_electricity_price,
  "replacement_infeasible_with_missing_city_electricity_price",
  replacement_status == "no_feasible_replacement",
  "replacement_infeasible_among_priced_candidates",
  default = NA_character_
)]

city_results[, c(
  "gw_cost_input_2019usd_m3",
  "gw_capacity_input_m3_year"
) := NULL]

stopifnot(all(city_results[
  urban_id %in% options$urban_id,
  has_valid_city_electricity_price
]))

nonempty_source_parts <- Filter(Negate(is.null), source_parts)

if (length(nonempty_source_parts)) {
  source_results <- rbindlist(
    nonempty_source_parts,
    use.names = TRUE,
    fill = TRUE
  )
} else {
  source_results <- data.table(
    urban_id = integer(),
    source_key = character(),
    source_type = character(),
    source_id = character(),
    design_id = integer(),
    max_delivery_m3_year = numeric(),
    fixed_cost_2019usd_year = numeric(),
    variable_cost_2019usd_m3 = numeric(),
    quantity_m3_year = numeric(),
    source_cost_2019usd_year = numeric()
  )
}

feasible <- city_results[
  replacement_status == "feasible_replacement"
]

stopifnot(
  nrow(city_results) == nrow(city),
  !anyDuplicated(city_results$urban_id),
  !anyDuplicated(source_results[, .(urban_id, source_key)]),
  all(feasible$extra_cost_2019usd_year >= 0),
  all(abs(
    feasible$replacement_surface_m3_year +
      feasible$replacement_groundwater_m3_year -
      feasible$demand_m3_year
  ) <= pmax(1e-3, 1e-6 * feasible$demand_m3_year)),
  all(is.na(city_results[
    replacement_status != "feasible_replacement",
    extra_cost_2019usd_year
  ])),
  all(source_results$urban_id %in% feasible$urban_id),
  all(!source_results[, paste(urban_id, source_key)] %in%
        chosen[, paste(urban_id, source_key)])
)

fwrite(
  city_results,
  file.path(
    analysis_dir,
    paste0("z_city_selected_portfolio_replacement_", scenario_tag, ".csv")
  )
)

fwrite(
  source_results,
  file.path(
    analysis_dir,
    paste0("z_city_replacement_sources_", scenario_tag, ".csv")
  )
)

print(city_results[, .N, by = replacement_status])
print(city_results[
  !is.na(missing_value_reason),
  .N,
  by = .(missing_value_reason, groundwater_input_status)
])
print(city_results[, .(
  valued_cities = sum(replacement_status == "feasible_replacement"),
  extra_cost_for_feasible_cities_2019usd_year =
    sum(extra_cost_2019usd_year, na.rm = TRUE)
)])
```
